\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.1cm]{geometry}
\usepackage{booktabs}
\usepackage[colorlinks=true, linkcolor=black, urlcolor=black, citecolor=black]{hyperref}
\setlength{\parskip}{0.35em}
\setlength{\parindent}{0em}

\begin{document}

\begin{center}
{\Large\bfseries Stem Agent for Real Python Bug Repair}\\
Andrea Gennari
\end{center}

\section{Goal and Scope}

The project asks whether a minimal agent can specialize itself for a class of
tasks. I instantiate the idea in a narrow, objectively scored domain: Python
bug repair on BugsInPy. The agent starts as a single-call direct bug fixer and
searches over a bounded configuration space: prompt constraints, reasoning
strategy, tool availability, examples, maximum iterations, and temperature.

The important design choice is that this is not a universal agent. It is a
stem agent for one task class: repairing localized Python bugs when given a
failing test output and candidate buggy line ranges. That scope is deliberately
smaller than SWE-bench-style issue resolution. It avoids synthetic benchmark
saturation while keeping the evaluation runnable in a one-day setting.

\section{Benchmark and Splits}

The benchmark is a frozen subset of 26 real BugsInPy bugs from \texttt{youtube-dl},
stored in \texttt{benchmark/subset.json}. Each task contains the project, bug id,
category, difficulty, and changed-line count. The harness checks out the buggy
commit, copies the maintainer test files from the fixed commit, applies the
agent's JSON edits, and runs the BugsInPy test command under WSL.

The prompt intentionally uses a fault-localization oracle: snippets around the
reference patch's old-file line ranges are shown, but the fixed code is never
shown. This means the benchmark evaluates patch generation and local reasoning,
not full repository-scale bug localization. That limitation is explicit because
otherwise the evaluation would overclaim.

The split is fixed in \texttt{benchmark/splits.json}:

\begin{center}
\begin{tabular}{lcl}
\toprule
Split & Size & Role \\
\midrule
Vital signs & 4 & Regression-safety invariant during differentiation \\
Calibration & 6 & Sensor analysis only \\
Dev & 8 & Champion/challenger selection \\
Test & 8 & Final held-out evaluation \\
\bottomrule
\end{tabular}
\end{center}

Vital-sign tasks were selected from bugs solved by the vanilla direct baseline.
Calibration, dev, and test are disjoint and fixed before any stem-agent search.
The test split is small, so all percentages are reported with counts.

\section{Stem-Agent Mechanism}

The agent genome is the \texttt{AgentConfig} dataclass in \texttt{src/config.py}.
It captures the system prompt, few-shot examples, output format, enabled tools,
tool policy, approach (\texttt{direct}, \texttt{cot}, \texttt{react},
\texttt{plan\_execute}), maximum iterations, temperature, and model. The
validator rejects malformed configurations, so every mutation produces a
runnable agent or no change.

The differentiation loop is implemented in \texttt{src/differentiator.py}.
Each iteration samples one or two mutation operators from \texttt{src/mutator.py},
evaluates the challenger on dev plus vital signs, and promotes it only if dev
pass@1 improves by more than \texttt{min\_improvement}. The safeguard rejects a
challenger if it regresses the vital-sign split. This is the main engineering
idea: specialization is allowed only when it does not break simple behavior the
base agent already had.

The sensor runs once on calibration results. In the one-day pipeline I added a
deterministic \texttt{--heuristic} mode to avoid spending extra calls only for
diagnostics. The generated report is in \texttt{results/sensor/sensor\_report.json}.
For this subset it found calibration pass@1 of 1/6, with failures concentrated
in \texttt{missing\_check}, \texttt{wrong\_variable}, and \texttt{other}. It
suggested \texttt{add\_constraint}, \texttt{specialize\_for\_category}, and
\texttt{switch\_approach}. The report explicitly records
\texttt{reference\_fixes\_used=false}; no fixed BugsInPy code is used by the
sensor.

\section{Current Evidence}

The available real-bug model-selection result is the stored
\texttt{vanilla\_direct} run on 25 BugsInPy bugs:

\begin{center}
\begin{tabular}{lcc}
\toprule
Model & Solved & pass@1 \\
\midrule
\texttt{gpt-5.4-nano-2026-03-17} & 6/25 & 24.0\% \\
\texttt{gpt-5.4-mini-2026-03-17} & 7/25 & 28.0\% \\
\bottomrule
\end{tabular}
\end{center}

The selected model is \texttt{gpt-5.4-mini-2026-03-17}. This is not a
statistically strong model comparison; it is a pragmatic one-day choice because
the gap is small and the sample is small.

On the frozen 8-task test split, the stored Phase 0 baseline solves 1/8:
\texttt{youtube-dl:13}. The other seven selected test bugs fail under
\texttt{vanilla\_direct}. The final after-number for \texttt{stem\_agent} is
produced by:

\begin{verbatim}
python scripts/run_differentiation.py --max-iterations 5
python scripts/run_final_eval.py --agents vanilla_direct stem_agent
\end{verbatim}

The code writes the final comparison to
\texttt{results/final\_eval/test\_results.json}, including pass@1, bootstrap
confidence intervals, cost, and McNemar's test when available. I do not claim a
stem-agent improvement unless that file contains it.

\section{Failure Modes and Tradeoffs}

The first attempted benchmark was synthetic and was abandoned because standard
direct prompting solved most tasks. That would have made any before/after claim
uninformative. Moving to BugsInPy made the task harder and exposed more realistic
failures: invalid JSON edits, syntactically invalid patches, patches that apply
but preserve the wrong behavior, and failures where the localized line range is
not enough context.

The main risk is complexity without measured gain. The implementation therefore
keeps the search space bounded and logs every accepted or rejected mutation. If
the stem agent does not beat \texttt{vanilla\_direct}, the correct conclusion is
not that self-specialization works; it is that this bounded search did not find
a better configuration within the budget.

The second limitation is the fault-localization oracle. Showing candidate line
ranges makes the task cheaper and isolates patch generation, but it also makes
the setting easier than full real-world bug repair. A stronger follow-up would
replace this with retrieval over the repository and score on a larger,
multi-project BugsInPy or SWE-bench Lite subset.

\section{What I Would Do Next}

With more time I would expand beyond \texttt{youtube-dl}, run the full
stem-agent search several times with different seeds, compare against
\texttt{vanilla\_cot} and a hand-tuned baseline, and report per-category
confidence intervals. I would also add a BugsInPy-aware verification tool for
candidate JSON edits, because the current ReAct tools were originally written
for the synthetic harness and are only partially useful on repository checkouts.

\begin{thebibliography}{9}

\bibitem{widyasari2020}
Ratnadira Widyasari et al.
\newblock BugsInPy: A database of existing bugs in Python programs to enable
controlled testing and debugging studies.
\newblock In \emph{Proceedings of ESEC/FSE}, 2020.

\bibitem{jimenez2024swebench}
Carlos E. Jimenez et al.
\newblock SWE-bench: Can Language Models Resolve Real-World GitHub Issues?
\newblock In \emph{Proceedings of ICLR}, 2024.

\end{thebibliography}

\end{document}
